---
title: "Final Project BIOS 611"
author: "Rebecca Jenkins"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction
The Oasis Longitudinal data set consists of 150 participants, aged 60 to 96, who received an MRI scan on two or more visits, each visit separated by at least a year. The cognitive status (demented/non demented) was documented throughout the study.

Note: This data was obtained from the Open Access Series of Imaging Studies (OASIS), which is comprised by data made available from the Washington University Alzheimer’s Disease Research Center, Dr. Randy Buckner at the Howard Hughes Medical Institute (HHMI) at Harvard University, the Neuroinformatics Research Group (NRG) at Washington University School of Medicine, and the Biomedical Informatics Research Network (BIRN).

```{r}
install.packages("kableExtra")
install.packages("knitr")
install.packages("ggplot2")
install.packages("gridExtra")
install.packages("grid")
install.packages("dplyr")
install.packages("tidyr")
install.packages("kableExtra")
install.packages("knitr")
```


LOADING LIBRARIES
```{r}
library(ggplot2)
library(gridExtra)
library(grid) 
library(dplyr)
library(tidyr)
library(kableExtra)
library(knitr)
```

IMPORTING DATA
```{r}
oasis_data <- read.csv("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/data/oasis_longitudinal.csv")
```

First, the demographics and collected data of study are shown below.

DEMOGRAPHICS FIGURE GENERATION CODE
```{r}
# AGE Variable calculations and figure generation
  # Calculations of age summary statistics
mean_age <- mean(oasis_data$Age, na.rm = TRUE)
median_age <- median(oasis_data$Age, na.rm = TRUE)
min_age <- min(oasis_data$Age, na.rm = TRUE)
q1_age <- quantile(oasis_data$Age, 0.25, na.rm = TRUE)
q3_age <- quantile(oasis_data$Age, 0.75, na.rm = TRUE)
max_age <- max(oasis_data$Age, na.rm = TRUE)
  # Data frame to hold the results
age_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_age, median_age, min_age, q1_age, q3_age, max_age)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = Age)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution with Summary Statistics",
       x = "Age (Years)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(age_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
age_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = 88, xmax = Inf, ymin = 0, ymax = 55)
  # Save RDS
saveRDS(age_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_stats.rds");


# EDUC Variable calculations and figure generation
  # Calculations of education summary statistics
mean_educ <- mean(oasis_data$EDUC, na.rm = TRUE)
median_educ <- median(oasis_data$EDUC, na.rm = TRUE)
min_educ <- min(oasis_data$EDUC, na.rm = TRUE)
q1_educ <- quantile(oasis_data$EDUC, 0.25, na.rm = TRUE)
q3_educ <- quantile(oasis_data$EDUC, 0.75, na.rm = TRUE)
max_educ <- max(oasis_data$EDUC, na.rm = TRUE)
  # Data frame to hold the results
educ_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_educ, median_educ, min_educ, q1_educ, q3_educ, max_educ)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = EDUC)) +
  geom_histogram(binwidth = 2, fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Years of Education Distribution with Summary Statistics",
       x = "Education (Years)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(educ_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
educ_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = 0, xmax = 15, ymin = 0, ymax = 190)
  # Save RDS
saveRDS(educ_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/educ_stats.rds");


# CDR (Clinical Dementia Rating) bar chart
  # Creating bar chart for CDR
cdr_plot <- ggplot(oasis_data, aes(x = CDR)) +
  geom_bar(fill = "lightgrey", color = "black", alpha = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.5, size = 3.5) +
  labs(title = "Clinical Dementia Rating (CDR)", 
       x = "Clinical Dementia Rating", 
       y = "Count") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.text.x = element_text(angle = 0, hjust = 0.5),    
  )
  # Save RDS
saveRDS(cdr_plot, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/cdr_plot.rds");


# M.F bar chart
  # Creating bar chart for CDR
sex_plot <- ggplot(oasis_data, aes(x = M.F)) +
  geom_bar(fill = "aquamarine2", color = "black", alpha = 0.7) + 
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.5, size = 3.5) + 
  labs(title = "Sex of Participants", 
       x = "Sex", 
       y = "Count") +
  scale_x_discrete(labels = c("M" = "Male", "F" = "Female")) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.text.x = element_text(angle = 0, hjust = 0.5),    
  )
  # Save RDS
saveRDS(sex_plot, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/sex_plot.rds");


# Hand bar chart
hand_count <- oasis_data %>%
  count(Hand) %>%
  mutate(Hand = factor(Hand, levels = c("R", "L"))) %>%
  complete(Hand, fill = list(n = 0))
  # Adding L level because there are no left handed participants
hand_plot <- ggplot(hand_count, aes(x = Hand, y = n)) +
  geom_bar(stat = "identity", fill = "darkolivegreen2", color = "black", alpha = 0.7) + 
  geom_text(aes(label = n), vjust = -0.5, size = 3.5) + 
  labs(title = "Dominant Hand of Participants", 
       x = "Dominant Hand", 
       y = "Count") +
  scale_x_discrete(labels = c("R" = "Right", "L" = "Left")) + 
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
  # Save RDS
saveRDS(hand_plot, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/hand_plot.rds");


# Group (Nondemeneted/Demented) bar chart
  # Creating bar chart for group
group_plot <- ggplot(oasis_data, aes(x = Group)) +
  geom_bar(fill = "azure3", color = "black", alpha = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.5, size = 3.5) +
  labs(title = "Dementia Status of Participants", 
       x = "Dementia Status", 
       y = "Count") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.text.x = element_text(angle = 0, hjust = 0.5),    
  )
  # Save RDS
saveRDS(group_plot, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/group_plot.rds");


# SES bar chart
oasis_data$alpha_value <- ifelse(oasis_data$SES == "Missing", 0.5, 0.7)
fill_colors <- c("Missing" = "deepskyblue", 
                 "1" = "deepskyblue",    
                 "2" = "deepskyblue",
                 "3" = "deepskyblue",
                 "4" = "deepskyblue",
                 "5" = "deepskyblue")

ses_plot <- ggplot(oasis_data, aes(x = SES)) +
  geom_bar(aes(fill = SES, alpha = alpha_value), color = "black") +  
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.5, size = 3.5) +
  labs(title = "Socioeconomic Status of Participants", 
       x = "Hollingshead Index of Social Position", 
       y = "Count") +
  scale_x_discrete(labels = c("1" = "1 (Highest Status)", "5" = "5 (Lowest Status)")) +
  scale_fill_manual(values = fill_colors) + 
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.text.x = element_text(angle = 0, hjust = 0.5),  
    legend.position = "none"
  )
  # Save RDS
saveRDS(ses_plot, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/ses_plot.rds");
```

DEMOGRAPHICS FIGURES
```{r}
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/educ_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/cdr_plot.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/sex_plot.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/hand_plot.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/group_plot.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/ses_plot.rds");
```

COLLECTED DATA FIGURE GENERATION CODE
```{r}
# MMSE variable calculations and figure generation
  # Remove rows with missing values
oasis_data_clean <- oasis_data %>% filter(!is.na(MMSE))
  # Create Normal/Not Normal category
oasis_data_clean$MMSE_category <- ifelse(oasis_data_clean$MMSE >= 25, "Normal", "Not Normal")
  # Calculate MMSE summary statistics 
mean_mmse <- mean(oasis_data_clean$MMSE, na.rm = TRUE)
median_mmse <- median(oasis_data_clean$MMSE, na.rm = TRUE)
min_mmse <- min(oasis_data_clean$MMSE, na.rm = TRUE)
q1_mmse <- quantile(oasis_data_clean$MMSE, 0.25, na.rm = TRUE)
q3_mmse <- quantile(oasis_data_clean$MMSE, 0.75, na.rm = TRUE)
max_mmse <- max(oasis_data_clean$MMSE, na.rm = TRUE)
  # Create data frame
mmse_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_mmse, median_mmse, min_mmse, q1_mmse, q3_mmse, max_mmse)
)
  # Histogram for MMSE 
p <- ggplot(oasis_data_clean, aes(x = MMSE, fill = MMSE_category)) +
  geom_histogram(binwidth = 2, color = "black", alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("Normal" = "paleturquoise", "Not Normal" = "lightseagreen")) + 
  labs(
    title = "Mini Mental State Examination Distribution with Summary Statistics",
    subtitle = "MMSE values classified as normal (≥ 25) and not normal (< 25)",  
    x = "MMSE",
    y = "Frequency",
    fill = "MMSE Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 10),
    legend.position = "right",
    plot.margin = margin(t = 10, r = 10, b = 100, l = 10) 
  )
  # Create table
table_grob <- tableGrob(mmse_stats, rows = NULL, theme = ttheme_minimal())
  # Combine histogram and table
mmse_stats <- p + annotation_custom(grob = table_grob, 
                                    xmin = 88, xmax = Inf, ymin = 0, ymax = 55) +
  annotation_custom(
    grob = textGrob("Two data points removed due to missing MMSE value.", 
                    x = 0.5, y = -0.1, gp = gpar(fontsize = 9, fontface = "italic")),
    xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = -Inf
  ) +
  coord_cartesian(clip = "off") +  # Prevent clipping of annotation
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 10),
    legend.position = "right",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)  # Keep margins small, but ensure plot area is large enough
  )
  # Save RDS
saveRDS(mmse_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_stats.rds");


# MR Delay variable calculations and figure generation
  # Calculations of education summary statistics
mean_mr <- mean(oasis_data$MR.Delay, na.rm = TRUE)
median_mr <- median(oasis_data$MR.Delay, na.rm = TRUE)
min_mr <- min(oasis_data$MR.Delay, na.rm = TRUE)
q1_mr <- quantile(oasis_data$MR.Delay, 0.25, na.rm = TRUE)
q3_mr <- quantile(oasis_data$MR.Delay, 0.75, na.rm = TRUE)
max_mr <- max(oasis_data$MR.Delay, na.rm = TRUE)
  # Data frame to hold the results
mr_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_mr, median_mr, min_mr, q1_mr, q3_mr, max_mr)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = MR.Delay)) +
  geom_histogram(binwidth = 100, fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "MR Delay Time (Contrast) Distribution with Summary Statistics",
       x = "Time (Days)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(mr_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
mr_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = 1250, xmax = Inf, ymin = 0, ymax = 190)
  # Save RDS
saveRDS(mr_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mr_stats.rds");


# eTIV Delay variable calculations and figure generation
  # Calculations of education summary statistics
mean_eTIV <- mean(oasis_data$eTIV, na.rm = TRUE)
median_eTIV <- median(oasis_data$eTIV, na.rm = TRUE)
min_eTIV <- min(oasis_data$eTIV, na.rm = TRUE)
q1_eTIV <- quantile(oasis_data$eTIV, 0.25, na.rm = TRUE)
q3_eTIV <- quantile(oasis_data$eTIV, 0.75, na.rm = TRUE)
max_eTIV <- max(oasis_data$eTIV, na.rm = TRUE)
  # Data frame to hold the results
eTIV_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_eTIV, median_eTIV, min_eTIV, q1_eTIV, q3_eTIV, max_eTIV)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = eTIV)) +
  geom_histogram(binwidth = 100, fill = "gray82", color = "black", alpha = 0.7) +
  labs(title = "Estimated Total Intracranial Volume Distribution with Summary Statistics",
       x = expression("eTIV ("* "cm"^3 * ")"), y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(eTIV_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
eTIV_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = 1700, xmax = Inf, ymin = 0, ymax = 130)
  # Save RDS
saveRDS(eTIV_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/eTIV_stats.rds");


# nWBV Delay variable calculations and figure generation
  # Calculations of education summary statistics
mean_nWBV <- mean(oasis_data$nWBV, na.rm = TRUE)
median_nWBV <- median(oasis_data$nWBV, na.rm = TRUE)
min_nWBV <- min(oasis_data$nWBV, na.rm = TRUE)
q1_nWBV <- quantile(oasis_data$nWBV, 0.25, na.rm = TRUE)
q3_nWBV <- quantile(oasis_data$nWBV, 0.75, na.rm = TRUE)
max_nWBV <- max(oasis_data$nWBV, na.rm = TRUE)
  # Data frame to hold the results
nWBV_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_nWBV, median_nWBV, min_nWBV, q1_nWBV, q3_nWBV, max_nWBV)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = nWBV)) +
  geom_histogram(binwidth = 0.02, fill = "greenyellow", color = "black", alpha = 0.7) +
  labs(title = "Normalize Whole Brain Volume Distribution with Summary Statistics",
       x = "Percent of Voxels Labeled as Gray/White Matter", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(nWBV_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
nWBV_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = Inf, xmax = 0.75, ymin = Inf, ymax = 35)
  # Save RDS
saveRDS(nWBV_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_stats.rds");


# ASF Delay variable calculations and figure generation
  # Calculations of ASF summary statistics
mean_ASF <- mean(oasis_data$ASF, na.rm = TRUE)
median_ASF <- median(oasis_data$ASF, na.rm = TRUE)
min_ASF <- min(oasis_data$ASF, na.rm = TRUE)
q1_ASF <- quantile(oasis_data$ASF, 0.25, na.rm = TRUE)
q3_ASF <- quantile(oasis_data$ASF, 0.75, na.rm = TRUE)
max_ASF <- max(oasis_data$ASF, na.rm = TRUE)
  # Data frame to hold the results
ASF_stats <- data.frame(
  Statistic = c("Mean", "Median", "Min", "Q1", "Q3", "Max"),
  Value = c(mean_ASF, median_ASF, min_ASF, q1_ASF, q3_ASF, max_ASF)
)
  # Histogram for age
p <- ggplot(oasis_data, aes(x = ASF)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue1", color = "black", alpha = 0.7) +
  labs(title = "Atlas Scaling Factor Distribution with Summary Statistics",
       x = "Atlas Scaling Factor (Unitless)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold") 
  )
  # Table of summary statistics
table_grob <- tableGrob(ASF_stats, rows = NULL, theme = ttheme_minimal())
  # Combining the histogram and table
ASF_stats <- p + annotation_custom(grob = table_grob, 
                     xmin = Inf, xmax = 1.35, ymin = Inf, ymax = 35)
  # Save RDS
saveRDS(ASF_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/ASF_stats.rds");
```

COLLECTED DATA
```{r}
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mr_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/eTIV_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/ASF_stats.rds");
```
# Analysis
Next, a few interesting connections will be analyzed.

AGE AND DEMENTIA GROUP
The first is age and group (demented/non-demented). Though these patients are 60 and older, it was expected that the demented group would be older than the non-demented group, especially because some patients changed groups from non-demented to demented over the course of the study. However, it was found that the non-demented and demented groups were not significantly different in age, but the converted group was significantly different than the demented group  with a p-value of 0.034, through the Tukey's HSD test.

SES AND EDUCATION
It was expected for years of education to decrease as Hollingshead Index of Social Position increases, because the status decreases with increasing Hollingshead Index of Social Position, where 1 is the highest status and 5 is the lowest. A linear regression was performed and as expected, the model was statistically significant with a p-value of 2.2e-16, showing that the predicted years of education decrease with decreasing socioeconomic status.

AGE AND CDR
It was expected for CDR to increase with age. The CDR is a 0-3 scale, with severe cognitive impairment being a 3. This dataset had CDR values ranging 0-2, with only 3 participants at a score of a 2. From the ANOVA with the Tukey's HSD test, there were no significant differences found in the age of each CDR rating. However, it would be interesting to redo this study and see if any differences arise when including more participants with a CDR of 2-3. 

GROUP AND MMSE 
It was expected that MMSE would be lower in the demented group because the MMSE assesses cognitive impairment, where lower scores indicate more significant cognitive impairment. Through an ANOVA with Tukey's HSD test, it was found that the demented group was significantly different than both the converted and non-demented group with a p-value of 0. An interesting finding from this analysis is that the converted and non-demented group are not significantly different. This may be explained by higher MMSE scores due to patients being in the early stages of dementia and cognitive impairment, or that the MMSE was performed while these patients were still considered non-demented.

nWBV AND AGE BY GROUP
nWBV was shown to decline significantly with increasing age in previous studies, so this was analyzed separately by dementia group. All three groups had a linear correlation between nWBV and age, which confirmed that nWBV decreased with increasing age. The rates of decline were slightly different. The nondemented group showed a rate of decline at 0.0028 per year, whereas the converted group showed a 35% higher rate of decline than the nondemented group (0.0038 per year). This may be due to the converted group being in a more active cognitive decline. Interestingly, the demented group showed a lower rate of decline than both groups at 0.0019 per year, which may be due to that group having a lower nWBV to start. An interesting future analysis would be to reanalyze this data by percentage instead of raw value, to see if a different result shows for the demented group. 

```{r}
# Age By Group 
  # Box Plot
age_group_box <- ggplot(oasis_data, aes(x = Group, y = Age, fill = Group)) + 
  geom_boxplot() +
  labs(title = "Box Plot of Age by Group", 
       x = "Group", 
       y = "Age") +
  scale_fill_manual(values = c("paleturquoise2", "olivedrab2", "gray80")) +
  theme_minimal() +
  theme(
    legend.position = "none",  
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)  
  )
  # ANOVA and Tukey's HSD
anova_result <- aov(Age ~ Group, data = oasis_data)
summary(anova_result)
tukey_result <- TukeyHSD(anova_result)
age_group_tukey_table <- as.data.frame(tukey_result$Group)
  # Save RDS
saveRDS(age_group_box, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_group_box.rds");
saveRDS(age_group_tukey_table, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_group_anova.rds");


# Education By SES
  # Linear Regression
    # Removing missing data
oasis_data_clean <- oasis_data[!is.na(oasis_data$SES) & !is.na(oasis_data$EDUC), ]
oasis_data_clean$SES <- as.numeric(as.character(oasis_data_clean$SES))
    # Model
model <- lm(EDUC ~ SES, data = oasis_data_clean)
summary(model)
intercept <- coef(model)[1]
slope <- coef(model)[2]
f_value <- summary(model)$fstatistic[1]
df1 <- summary(model)$fstatistic[2]
df2 <- summary(model)$fstatistic[3]
p_value <- pf(f_value, df1, df2, lower.tail = FALSE)
    # Plot labels
eq_label <- paste("y = ", round(intercept, 2), " + ", round(slope, 2), " * x")
stat_label <- paste("F = ", round(f_value, 2), ", p = ", format.pval(p_value, digits = 4))
    # Plot
educ_ses_reg <- ggplot(oasis_data_clean, aes(x = SES, y = EDUC, color = factor(SES))) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black") + 
  labs(title = "Linear Regression of Years of Education vs. Socioeconomic Status",
       x = "Hollingshead Index of Social Position",
       y = "Years of Education",
       color = "SES") +
  scale_x_discrete(labels = c("1" = "1 (Highest Status)", "5" = "5 (Lowest Status)")) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14)) +  
  annotate("text", x = max(oasis_data_clean$SES) - 1, y = max(oasis_data_clean$EDUC) - 3, 
           label = eq_label, size = 4, color = "black", hjust = 0, vjust = 0) +
  annotate("text", x = max(oasis_data_clean$SES) - 1, y = max(oasis_data_clean$EDUC) - 5, 
           label = stat_label, size = 4, color = "black", hjust = 0, vjust = 0)
  # Save RDS
saveRDS(educ_ses_reg, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/educ_ses_reg.rds");


# Age by CDR
  # Mean and Count Table
age_cdr_stats <- oasis_data %>%
  group_by(CDR) %>%
  summarise(
    mean_age = mean(Age, na.rm = TRUE),  
    count = n()                       
  )
  # Box Plot
oasis_data$CDR <- as.factor(oasis_data$CDR)
age_cdr_box <- ggplot(oasis_data, aes(x = CDR, y = Age, fill = CDR)) + 
  geom_boxplot() +
  labs(title = "Box Plot of Age by Clinical Dementia Rating", 
       x = "Clinical Dementia Rating", 
       y = "Age") +
  scale_fill_manual(values = c("olivedrab1", "olivedrab2", "olivedrab3", "olivedrab4")) +
  theme_minimal() +
  theme(
    legend.position = "none",  
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)  
  )
  # ANOVA and Tukey's HSD
anova_result <- aov(Age ~ CDR, data = oasis_data)
summary(anova_result)
tukey_result <- TukeyHSD(anova_result)
age_cdr_tukey_table <- as.data.frame(tukey_result$CDR)
  # Save RDS
saveRDS(age_cdr_stats, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_stats.rds");
saveRDS(age_cdr_box, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_box.rds");
saveRDS(age_cdr_tukey_table, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_anova.rds");


# MMSE By Group
  # Data Cleaning
oasis_data_clean <- oasis_data %>%
  filter(!is.na(MMSE), !is.infinite(MMSE), !is.na(Group), !is.infinite(Group))
  # Box Plot
mmse_group_box <- ggplot(oasis_data_clean, aes(x = Group, y = MMSE, fill = Group)) + 
  geom_boxplot() +
  labs(title = "Box Plot of Mini Mental State Examination (MMSE) By Group", 
       x = "Group", 
       y = "MMSE") +
  scale_fill_manual(values = c("paleturquoise2", "olivedrab2", "gray80")) +
  theme_minimal() +
  theme(
    legend.position = "none",  
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)  
  ) +
  geom_hline(yintercept = 25, linetype = "dashed", color = "red", linewidth = 1) +
  # Add shaded region above Y = 25
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 25, ymax = Inf), 
            fill = "lightblue", alpha = 0.01) +
  # Add label for 'Normal Range'
  annotate("text", x = 1.5, y = 30, label = "Normal Range", size = 4, 
           color = "black", fontface = "italic", hjust = -3.25, vjust = 5)
  # ANOVA and Tukey's HSD
mmse_anova_result <- aov(MMSE ~ Group, data = oasis_data_clean)
summary(mmse_anova_result)
mmse_tukey_result <- TukeyHSD(mmse_anova_result)
mmse_tukey_result
mmse_group_tukey_table <- as.data.frame(mmse_tukey_result$Group)
  # Save RDS
saveRDS(mmse_group_box, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_group_box.rds");
saveRDS(mmse_group_tukey_table, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_group_anova.rds");


# nWBV and Age By Dementia Group
  # Combined plot
intercept_non_demented <- format(coef(model_non_demented)[1], scientific = FALSE, digits = 3)
slope_non_demented <- format(coef(model_non_demented)[2], scientific = TRUE, digits = 3)

intercept_demented <- format(coef(model_demented)[1], scientific = FALSE, digits = 3)
slope_demented <- format(coef(model_demented)[2], scientific = TRUE, digits = 3)

intercept_converted <- format(coef(model_converted)[1], scientific = FALSE, digits = 3)
slope_converted <- format(coef(model_converted)[2], scientific = TRUE, digits = 3)

oasis_data_clean2 <- oasis_data %>%
  filter(!is.na(nWBV), !is.infinite(nWBV), !is.na(Group), !is.infinite(Group), !is.na(Age), !is.infinite(Age))

nWBV_age_group_reg <- ggplot(oasis_data_clean2, aes(x = Age, y = nWBV, color = Group)) +
  geom_point(size = 2, alpha = 0.6) +  
  geom_smooth(method = "lm", se = FALSE, aes(color = Group), linetype = "solid") + 
  
  annotate("text", x = 89.17, y = 0.85, label = paste("Converted: y =", intercept_converted, 
                                                 "+", slope_converted, "* x"), size = 4, color = "black") +
  annotate("text", x = 89.5, y = 0.83, label = paste("Demented: y =", intercept_demented, 
                                                 "+", slope_demented, "* x"), size = 4, color = "black") +
  annotate("text", x = 90, y = 0.81, label = paste("Nondemented: y =", intercept_non_demented, 
                                                 "+", slope_non_demented, "* x"), size = 4, color = "black") +
  labs(title = "Linear Regression for Normalized Whole Brain Volume (nWBV) by Age and Group",
       x = "Age",
       y = "nWBV",
       color = "Group") + 
  scale_color_manual(values = c("Nondemented" = "firebrick", "Demented" = "dodgerblue3", "Converted" = "green4")) + 
  theme_minimal() +  
  theme(legend.position = "top",  
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12)) +
  xlim(60, 98) +  
  ylim(0.62, 0.85) 
  # Results table
slope_non_demented <- coef(model_non_demented)[2]
intercept_non_demented <- coef(model_non_demented)[1]
f_stat_non_demented <- summary(model_non_demented)$fstatistic[1]
p_value_non_demented <- summary(model_non_demented)$coefficients[2, 4]
slope_demented <- coef(model_demented)[2]
intercept_demented <- coef(model_demented)[1]
f_stat_demented <- summary(model_demented)$fstatistic[1]
p_value_demented <- summary(model_demented)$coefficients[2, 4]
slope_converted <- coef(model_converted)[2]
intercept_converted <- coef(model_converted)[1]
f_stat_converted <- summary(model_converted)$fstatistic[1]
p_value_converted <- summary(model_converted)$coefficients[2, 4]
regression_results <- data.frame(
  Group = c("Non-demented", "Demented", "Converted"),
  Intercept = c(intercept_non_demented, intercept_demented, intercept_converted),
  Slope = c(slope_non_demented, slope_demented, slope_converted),
  F_Statistic = c(f_stat_non_demented, f_stat_demented, f_stat_converted),
  P_Value = c(p_value_non_demented, p_value_demented, p_value_converted)
)
# Save RDS
saveRDS(nWBV_age_group_reg, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_age_group_reg.rds");
saveRDS(regression_results, file="/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_age_group_regression_results.rds");

```
```{r}
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_group_box.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_group_anova.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/educ_ses_reg.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_box.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_anova.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/age_cdr_stats.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_group_box.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/mmse_group_anova.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_age_group_reg.rds");
readRDS("/Users/rebeccajenkins/BIOS611/RJENKI_BIOS611Project/figures/nWBV_age_group_regression_results.rds");
```

# Conclusion

Significant Results
The age of the converted group was significantly different than the age of the demented group with a p-value of 0.034, but no other comparisons from that analysis were significant. Years of education do in fact decrease as Hollingshead Index of Social Position increases, as the model obtained through linear regression was statistically significant with a p-value of 2.2e-16. There were no significant differences in age by clinical dementia rating, but it should be noted that there was an underrepresentation of patients with a CDR of 2-3. It was found that the MMSE score of the demented group was significantly different than both the converted and non-demented group with a p-value of 0. The converted and non-demented group are not significantly different in terms of MMSE score. All three groups (nondemented, demented, converted) had a negative linear correlation between nWBV and age, which confirmed that nWBV decreased with increasing age. Of the three groups, the converted group showed a the largest rate of decline, and the demented group showed the smallest rate of decline.

Future Directions
It would be interesting to redo this study and see if any differences in age arise when including more participants with a CDR of 2-3. The normalized whole brain volume data vs. age analysis by group should be redone by using a percentage or normalizing the y-axis due to the demented group's lower starting nWBV to see if there is a larger rate of decline. 